---
title: "Classification Coding Challenge"
author: "Philip Waggoner, MACS 30100 <br /> University of Chicago"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE)
```

# Code Challenge

**To submit your solution, simply email me your final figure: <pdwaggoner@uchicago.edu>**
  
## Some basics:

  - *The first person to submit wins the 5% bonus on the current problem set* 
  
  - *Everyone* must submit their *attempt* at the challenge, whether completed or not, to today's normal code submission link on Canvas to receive credit. Normal grading guidelines apply.
  
  - All preprocessing is done; just load the data and begin.
  
  - I have answered questions 3 and 4 for you. But be careful to ensure my code integrates with your code. Keep names and objects well-organized.
  
  - Deliverable is a *single plot*, per the task description below.

  - You may use either R or Python, and you may use any approach within these languages to answer the questions.
  
  - Work independently. 
  
  - I am here to answer questions via the chat (*directly*).

## The Task

Replicate the figure from slide 17 in today's lecture notes, but for logistic regression versus kNN classification. That is: **plot a comparison of the test error over many fits of a kNN classifier (over a range of $k$) to the test error from a logistic regression classifier**. 

*Reminder*: The clock for the challenge runs out at 9:20 am CDT. At that point, just submit your attempt by tomorrow at 5 pm CDT to Canvas as normal.

## The Questions

1. Load the titanic data (`titanic.csv`), which has been cleaned and preprocessed for you including only relevant features for this challenge. 

2. Create a train/test split, at 70/30 respectively.

3. Fit *100* versions of a kNN classifier to the training data, with each fit corresponding to a different value of $k \in \{1, \dots, 100\}$. 

```{r}
mse_knn <- tibble(k = 1:100,
                  knn_train = map(k, ~ class::knn(dplyr::select(titanic_train, -Survived),
                                                  test = dplyr::select(titanic_train, -Survived),
                                                  cl = titanic_train$Survived, k = .)),
                  knn_test = map(k, ~ class::knn(dplyr::select(titanic_train, -Survived),
                                                 test = dplyr::select(titanic_test, -Survived),
                                                 cl = titanic_train$Survived, k = .)),
                  err_train = map_dbl(knn_train, ~ mean(titanic_test$Survived != .)),
                  err_test = map_dbl(knn_test, ~ mean(titanic_test$Survived != .)))
```

4. Record the testing error for *each fit* of the kNN classifier using the testing (30%) set.

```{r}
# done in the previous chunk responding to question 3
```

5. Fit a logistic regression to the training data, predicting the probability of survival (`Survived`) as a function of all other features in the data.

6. Record the testing error for the logistic regression using the testing (30%) set.

7. Plot the test error rate from all fits of the kNN classifier as line plot (with values of $k$ ranging along the X axis, and the range of testing error on the Y axis). Place a horizontal reference line on the plot showing the test error from the logistic regression. *Note*: The final plot created in this question is the only thing to be submitted/emailed to me. 
