---
title: "Classification, pt. 2"
author: "Philip Waggoner, MACS 30100 <br /> University of Chicago"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE)
```

# Some Changes

1. *On your own code*: 
    
  - Scale this portion back/not every class 
  - Complete on your own outside of class
  
2. *Class administration mode*: Giving asynchronous a try next week, week 5 on non-linear estimation:
  
	- More detailed notes (in place of a lecture) to accompany your required reading
	- More detailed code (in place of in-person sessions) to allow you to flexibly work through some sample code
	- There will be *no* on your own code section/code submission
	- There will be *no* Q&A requirement
	- There *will* be a normal weekly quiz to complete

3. *Q&A Requirement*: No more Q&A requirement, but still use many outlets available for questions as they arise

# Overview

Goals today: 

  - Cover basic model-based code in R and Python to set you up for your problem set this week
  - Coding challenge at the end, working independently -- winner gets extra 5% on problem set

# Classification with kNN, LDA, and QDA 

The data we are working with for this brief session is `Court.dta`, containing several features on the U.S. Supreme Court. Each row is a case, each column is a feature. The outcome of interest is `propetit`, which is 1 when the US Supreme Court voted in favor of the petitioner (the person/party who brought the case), and 0 if they voted against the petitioner. The other features are:

  - `liberal`: liberalism
  - `usparty`: US government a party to the case (0/1)
  - `ineffcou`: ineffective counsel (0/1)
  - `tcterror`: trial error (0/1)
  - `multpet`: multiple parties on the case (0/1)

**Goal**: predict supreme court ruling for (1) or against (0) petitioner.

```{r}
library(tidyverse)
library(here) # for loading data; this is *optional*
library(tidymodels) # for accuracy, splitting, etc.
library(foreign) # for the (stata) data
library(class) # for knn()
library(MASS) # for lda() and qda()

# read in court data
data <- read.dta(here("data", "Court.dta")) %>% 
  as_tibble() %>% 
  mutate(liberal = round(liberal * 100, 2))

# split into training/test set
split <- initial_split(data, 
                       prop = 0.7)

train <- training(split)
test <- testing(split)
```

## kNN 

The `knn()` function is from `class` package. The `tidymodels` ecosystem is more thorough, but also more complex, by comparison. Take a look at their helper page if you're interested in giving this a try for your problem set: <https://parsnip.tidymodels.org/reference/nearest_neighbor.html>.

```{r}
# fit the model using knn() from the class package
knn_mod <- knn(train, # training set
               test, # testing set
               train$propetit, # outcome from the training set
               k = 1) # local classifier, where k = 1 (feel free to change k to see differences in classification ability, i.e., comparing global vs. local behavior of the kNN classifier)

table(knn_mod, test$propetit) # create simple confusion matrix

mean(knn_mod == test$propetit) # calculate the overall accuracy rate
```

A few tips for Python:

Consider the `KNeighborsClassifier()` function from the `neighbors` submodule of `sklearn`. Also consider `confusion_matrix` and `classification_report` from `sklearn.metrics`.

## LDA 

Now, let's fit LDA to the data, with the same task/goal.

Now, we will fit an LDA classifier using the `lda()` function from the `MASS` package.

```{r}
# fit the LDA classifier
lda_mod <- lda(propetit ~ ., # same model syntax we have seen before: DV ~ on IV(s)
               data = train)

# inspect
lda_mod
```

Three key outputs: 

  - "Prior probabilities of groups"
  - "Group means"
  - "Coefficients of linear discriminants"

Now, make some predictions using `predict()`.

```{r}
# create a out of sample data frame of four observations/cases: 2 with a liberal (65) court and 2 with a conservative (35) court; 2 with the government as a party to the case and 2 without, for each liberal and conservative version of the court, holding other categorical features at median levels

df <- tibble(
  liberal = rep(c(35, 65), 2), 
  usparty = c(0, 0, 1, 1),
  ineffcou = rep(median(train$ineffcou), 4), # rep just means repeat the value x times, which in this case is 4, as we have 4 cases we're creating
  tcterror = rep(median(train$tcterror), 4),
  multpet = rep(median(train$multpet), 4))

# inspect the df object once you've made it so it makes more sense how things are constructed
df

# now predict
df_pred <- predict(lda_mod, df)
df_pred
```

A list with three elements:

  - `class`: classification assignments
  - `posterior`: posterior probability of 1 relative to a 0
  - `x`: linear discriminant values

Exploring the (default) threshold:

```{r}
# number of no votes
sum(df_pred$posterior[, 1] >= 0.5)
table(df_pred$posterior[, 1] >= 0.5)

# number of yes votes
sum(df_pred$posterior[, 2] > 0.5)
```

Change the threshold.

```{r}
sum(df_pred$posterior[, 2] > 0.4)
```

## QDA 

Now, we will use `qda()` from `MASS`. Most the same, except no coefficients, because the QDA classifier involves a *quadratic* (not linear as in LDA) combination of the predictors.

```{r}
qda_mod <- qda(propetit ~ ., 
               data = train)
```

Make predictions in the same way, with the same `df`.

```{r}
predict(qda_mod, df)
```

Not much change, just strong prediction results.

Some Python tips: 

Consider the `LinearDiscriminantAnalysis()`, `QuadraticDiscriminantAnalysis()` functions from the `discriminant_analysis` module of `sklearn`, and also the `confusion_matrix()`, `classification_report()`, and `precision_score()` functions from `sklearn.metrics` from `sklearn`. 

Example of a simple Python application of LDA using the `titanic.csv`, which you will use for the challenge next.

```{python}
import pandas as pd
import numpy as np

from sklearn.discriminant_analysis import LinearDiscriminantAnalysis, QuadraticDiscriminantAnalysis
from sklearn.metrics import confusion_matrix, classification_report, precision_score

titanic = pd.read_csv('titanic.csv')
titanic.head()

# define train/test split, e.g., 80/20
X_train = titanic[:80][['Pclass', 'Age', 'SibSp', 'Parch', 'Fare', 'Female']]
y_train = titanic[:80]['Survived']

X_test = titanic[80:][['Pclass', 'Age', 'SibSp', 'Parch', 'Fare', 'Female']]
y_test = titanic[80:]['Survived']

# fit QDA and inspect output
lda = LinearDiscriminantAnalysis()
model = lda.fit(X_train, y_train)

print(model.priors_)
print(model.coef_)

# evaluate/test
lda_preds = model.predict(X_test)

# inspect
print(confusion_matrix(lda_preds, y_test))
print(classification_report(y_test, lda_preds))
```

# Code Challenge

**To submit your solution, simply email me your final figure: <pdwaggoner@uchicago.edu>**
  
## Some basics:

  - *The first person to submit wins the 5% bonus on the current problem set* 
  
  - *Everyone* must submit their *attempt* at the challenge, whether completed or not, to today's normal code submission link on Canvas to receive credit. Normal grading guidelines apply.
  
  - All preprocessing is done; just load the data and begin.
  
  - I have answered questions 3 and 4 for you. But be careful to ensure my code integrates with your code. Keep names and objects well-organized.
  
  - Deliverable is a *single plot*, per the task description below.

  - You may use either R or Python, and you may use any approach within these languages to answer the questions.
  
  - Work independently. 
  
  - I am here to answer questions via the chat (*directly*).

## The Task

Replicate the figure from slide 17 in today's lecture notes, but for logistic regression versus kNN. That is: **plot a comparison of the test error over many fits of a kNN classifier (over a range of $k$) to the test error from a logistic regression classifier**. 

See `Code Challenge.Rmd` for challenge questions. The clock for the challenge runs out at 9:20 am CDT. At that point, just submit your attempt by tomorrow at 5 pm CDT to Canvas as normal.
